False Discovery Rates, A New Deal

### 2015-10-14 talks/twentymin_seminar.Rmd
updated seminar for CSH

## How can we make more progress together?

- Abstract our problems
- Develop benchmarks
- Build generic tools

@@

Goal: generic tools for shrinkage estimation

@@

- Analogously, one can assume that all Z scores near 0 are null. Efron refers to this as the ``Zero Assumption".

@@

```{r, echo=FALSE, include=FALSE}
require(fdrtool)
  hh.fdrtool = fdrtool(hh.pval,statistic="pvalue",plot=FALSE)
require(locfdr)
  hh.locfdr = locfdr(hh.zscore,nulltype=0,plot=0)
require(mixfdr)
  hh.mixfdr = mixFdr(hh.zscore,noiseSD=1,theonull=TRUE,plot=FALSE)
```

@@

  nullalthist(hh.zscore,hh.fdrtool$lfdr)  
```

@@

- In some cases the measurement precisions differ among units

- Eg effect sizes of rare SNPs have larger standard error than those of common SNPs

- Eg Expression levels of low-expressed genes have less precision than high-expressed genes

- If some effects are measured less precisely than others, those tests ``lack power"
and dilute signal, increasing FDR

@@

- Following previous work (e.g. Newton, Efron, Muralidharan) we take an empirical Bayes approach to FDR.

@@

- Various semi-parametric approaches taken to estimating $f_1$. For example,
Efron uses Poisson regression; Muralidharan uses mixture of normal distributions.

- Once $f_1$ and $\pi_0$ estimated, FDR calculations are straightforward.

@@

(cf Bayes inference from summary statistics; V.Johnson, J.Wakefield)

@@

- Estimating $g$ comes down to estimating $\pi$. Joint estimation of $\pi_0,\pi$ easy by maximum likelihood (EM algorithm).

- By allowing $K$ large, and $\sigma_k$ to span a dense grid of values,
we get a flexible unimodal symmetric distribution.

- Can approximate, arbitrarily closely, any scale mixture of normals.
Includes almost all priors used for sparse regression problems (spike-and-slab, double exponential/Laplace/Bayesian Lasso, horseshoe).

- Alternatively, a mixture of uniforms, with 0 as one end-point of the range,
provides still more flexibility, and in particular allows for asymmetry.

- If allow a very large number of uniforms this provides the non-parametric mle for $g$; cf Grenander 1953; Cordy + Thomas 1997.

@@

- This approach actually provides a full posterior distribution for each $\beta_j$.

- So easy to obtain point estimates and credible intervals.

- Because $g(\beta)$ is unimodal, the point estimates (and CIs) will tend to be ``shrunk" towards the overall mean (0).

@@

- So we call the approach ``Adaptive Shrinkage" (ASH).

## Recall Problem 1: distribution of alternative Z values multimodal
```{r, echo=FALSE,fig.height=4,fig.cap=""}
nullalthist(hh.zscore,hh.fdrtool$lfdr,main="qvalue")  
```

@@

## Some more extensive simulations

@@

All methods are, generally, "conservative" (deliberately).

@@

## A new problem: an embarrassment of riches

- If the null is mostly false, this approach can  provide unsettling results

- The FDR can be small for all observations, even those with $p \approx 1$!

- In the illustrative example, the maximum $q$ value is `r round(max(hh.ash$qvalue),2)`

## Perhaps we didn't really understand the question?

- Problem arises only if we insist on asking question "is $\beta_j=0$?"

- Given enough signal, we become convinced that very few of the $\beta_j=0$

- But for some $\beta_j$ we still may have little information about actual value

- Suggests a change of focus: ask for which $\beta_j$ are we confident about the sign (cf Gelman et al, 2012; Tukey 1961).

@@

## Generic procedure: embellishments

$$\hat{\beta}_j | \beta_j, s_j \sim t(\beta_j,s_j)$$
$$\beta_j | s_j \sim s_j^\alpha g(), \text{unimodal}.$$

@@

- Incorporate shrinkage of variances and not just means. (e.g. ``moderated $t$ test", Smyth et al).

@@

- NHGRI, Moore Foundation for funding.

>>>>

@@

Goal: generic tools for FDR and shrinkage estimation

@@

- Analogously, one can assume that all Z scores near 0 are null. Efron refers to this as the "Zero Assumption".

@@

```{r, echo=FALSE, include=FALSE}
#require(fdrtool)
#  hh.fdrtool = fdrtool(hh.pval,statistic="pvalue",plot=FALSE)
require(locfdr)
  hh.locfdr = locfdr(hh.zscore,nulltype=0,plot=0)
require(mixfdr)
  hh.mixfdr = mixFdr(hh.zscore,noiseSD=1,theonull=TRUE,plot=FALSE)
```

@@

  nullalthist(hh.zscore,hh.q$lfdr)  
```

@@

- Measurement precision can differ among units

    - Eg effects of rare SNPs have larger standard error than common SNPs

    - Eg low-expressed genes less precise than high-expressed genes

- Low precision measurements add noise, dilute signal, increase FDR

@@

- Following previous work (e.g. Newton, Efron, Muralidharan) we take an empirical Bayes approach.

@@

- Once $f_1$ and $\pi_0$ estimated, FDR calculations straightforward.

@@

@@

- Fix $K$ large, $\sigma_k$ fixed dense grid, estimate $\pi$.

@@

- Approach provides not only FDR estimates, but posterior distribution for each $\beta_j$.

- Because $g(\beta)$ is unimodal, the point estimates (and CIs) will tend to be "shrunk" towards the overall mean (0).

@@

- So we call the approach "Adaptive Shrinkage" (ASH).

## Recall Problem 1: distribution of alternative Z values multimodal
```{r, echo=FALSE,fig.height=4,fig.cap=""}
nullalthist(hh.zscore,hh.q$lfdr,main="qvalue")  
```

@@

## More extensive simulations

@@

- All methods are, generally, "conservative" (deliberately).

- `ash` generally least conservative.

@@

@@

@@

- Incorporate shrinkage of variances and not just means. (e.g. "moderated $t$ test", Smyth et al).

@@

- NHGRI, and the Gordon and Betty Moore Foundation for funding.
